author: "Trevor Eakes"
date: "`r Sys.Date()`"
title: "PCA and PCR proximate nutrient analysis and graphing"
output: html_document
params:
  plot_dir:    "output_all/plots"
  data_dir:    "output_all/export_data"
  report_dir:  "output_all/reports"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load script and directories
```{r}
# Source script from top-level project folder
source("../process_plate_run.R")

#— 1. pull from params (or use defaults)
plot_dir   <-  "output_all/plots"
data_dir   <-  "output_all/export_data"
report_dir <-  "output_all/reports"

#— 2. create all dirs
walk(
  c(plot = plot_dir, data = data_dir, reports = report_dir),
  ~ if (!dir_exists(.x)) dir_create(.x, recurse = TRUE)
)

#— 3. register plot_dir with knitr
opts_chunk$set(fig.path = paste0(plot_dir, "/"))
#
```

## Load input data
```{r load-excel-files, warning=FALSE}
# 1. Define the folder containing all input files
input_folder <- "input_all"

# 2. List all Excel and CSV files (full paths), excluding temp Excel files (~$)
all_files <- list.files(
  path = input_folder,
  pattern = "\\.(xlsx|csv)$",
  full.names = TRUE
)
all_files <- all_files[!grepl("^~\\$", basename(all_files))]

# Ensure `readxl` and `readr` are loaded if not already in your setup chunk
# library(readxl)
# library(readr)
# No need for library(fs) with this change!

# 3. Loop over each file
for (file in all_files) {
  # 3a. Create a clean object name (strip out folder & extension)
  # CHANGED: Using tools::file_path_sans_ext from base R
  name <- tools::file_path_sans_ext(basename(file))

  # 3b. Determine file type
  if (grepl("\\.xlsx$", file)) {
    # Read Excel file (2nd sheet if available, else 1st)
    sheet_names <- readxl::excel_sheets(file)
    sheet_to_read <- if (length(sheet_names) >= 2) sheet_names[2] else sheet_names[1]
    data <- suppressMessages(readxl::read_excel(file, sheet = sheet_to_read))
    message("Loaded Excel file: ", name, " (sheet = '", sheet_to_read, "')")
  } else if (grepl("\\.csv$", file)) {
    # Read CSV file with explicit UTF-8 encoding
    data <- suppressMessages(readr::read_csv(file,
                                             show_col_types = FALSE,
                                             locale = readr::locale(encoding = "UTF-8")))
    message("Loaded CSV file: ", name)
  }

  # 3c. Assign the data.frame to the global environment under "name"
  assign(name, data, envir = .GlobalEnv)
}

# --- IMPORTANT AFTER LOADING ALL FILES ---
# Even with explicit encoding, sometimes files can have truly malformed UTF-8
# or be in a different encoding altogether (e.g., Latin-1).
# If you still get UTF-8 errors *after* this, consider these troubleshooting steps:
# 1. Try a different encoding for CSVs: `locale = readr::locale(encoding = "Latin1")`
#    or `locale = readr::locale(encoding = "ISO-8859-1")` (Windows-1252 is often similar to Latin1)
# 2. After loading, apply the `iconv` fix to your merged data:
#    (Example for full_merged_data after it's created)
#    char_cols <- names(full_merged_data)[sapply(full_merged_data, is.character)]
#    for (col_name in char_cols) {
#      full_merged_data[[col_name]] <- iconv(full_merged_data[[col_name]], from = "", to = "UTF-8", sub = "byte")
#    }
#    This `iconv` step attempts to clean up any remaining invalid characters.
```
## Make the factors and color levels
```{r}
# For Life_S

#Code to collapse factors
desired_location_order <- c("Mendieta", "Concession", "Caro Caido", "7H", "Powder Ilo", "Ilo mariculture")
lifestage_levels           <- c("Carp", "Gam", "Gam/Tetra", "Tetra", "Unknown")

# Harmonized Colors
group_colors1 <- c(
  # Location colors
  "Mendieta"        = "#66c2a5",  # muted green
  "Concession"      = "#66e2a5",  # same as Mendieta
  "Caro Caido"      = "#fc8d62",  # coral orange
  "7H"              = "#8da0cb",  # soft blue
  "Powder Ilo"      = "#e78ac3",  # pink
  "Ilo mariculture" = "#a6d854",  # lime green

  # Life_S colors
  "Carp"       = "#ffd92f",  # yellow
  "Gam"        = "#e5c494",  # tan
  "Gam/Tetra"  = "#b3f3b3",  # gray
  "Tetra"      = "#a6cee3",  # light blue
  "Unknown"    = "#cab2d6"   # lavender
)
library(forcats)
pe_rep_final$Life_S <- fct_collapse(pe_rep_final$Life_S, "Gam/Tetra" = c("Gam", "Tetra"))
pe_rep_final$Life_S <- factor(pe_rep_final$Life_S, levels = lifestage_levels)

pr_rep_final$Life_S <- fct_collapse(pr_rep_final$Life_S, "Gam/Tetra" = c("Gam", "Tetra"))
pr_rep_final$Life_S <- factor(pr_rep_final$Life_S, levels = c("Carp", "Gam/Tetra", "Unknown"))

carb_rep_final$Life_S <- fct_collapse(carb_rep_final$Life_S, "Gam/Tetra" = c("Gam", "Tetra"))
carb_rep_final$Life_S <- factor(carb_rep_final$Life_S, levels = c("Carp", "Gam/Tetra", "Unknown"))

car_rep_final$Life_S <- fct_collapse(car_rep_final$Life_S, "Gam/Tetra" = c("Gam", "Tetra"))
car_rep_final$Life_S <- factor(car_rep_final$Life_S, levels = c("Carp", "Gam/Tetra", "Unknown"))

ash_rep_final$Life_S <- fct_collapse(ash_rep_final$Life_S, "Gam/Tetra" = c("Gam", "Tetra"))
ash_rep_final$Life_S <- factor(ash_rep_final$Life_S, levels = c("Carp", "Gam/Tetra", "Unknown"))


# repeat as needed

# For Location (example)
desired_location_order <- c("Mendieta", "Concession", "Caro Caido", "7H", "Powder Ilo", "Ilo mariculture"  )  # or whatever order you want

pr_rep_final$Location <- factor(pr_rep_final$Location, levels = desired_location_order)
carb_rep_final$Location <- factor(carb_rep_final$Location, levels = desired_location_order)
ash_rep_final$Location <- factor(ash_rep_final$Location, levels = desired_location_order)
car_rep_final$Location <- factor(car_rep_final$Location, levels = desired_location_order)
pe_rep_final$Location <- factor(pe_rep_final$Location, levels = desired_location_order)

# Define factor levels and color palette
  # or locations

```

## Make the plots
```{r}
########################## subset protein
# Protein
pr_location <- pr_rep_final %>%
  filter(!Location %in% c("Lima Market Freeze Dry", "Ilo Freeze Dry", "Ilo oven dry", "Ilo Fresh", "Lima Market Fresh"))

pr_location_cham <- pr_rep_final %>%
  filter(!variety %in% c("F.Glom"))

########################## Subset PE
pe_location <- pe_rep_final %>%
  filter(!Location %in% c("Lima Market Freeze Dry", "Ilo Freeze Dry", "Ilo oven dry", "Ilo Fresh", "Lima Market Fresh"))

pe_location_cham <- pe_location %>%
  filter(!variety %in% c("F.Glom"))
# 6. Run group comparisons and print outputs


# Apply color palette
group_colors <- c(
  # Location colors
  "Mendieta"        = "#66c2a5",
  "Concession"      = "#66c2a5",
  "Caro Caido"      = "#fc8d62",
  "7H"              = "#8da0cb",
  "Powder Ilo"      = "#e78ac3",
  "Ilo mariculture" = "#a6d854",
  # Life_S colors
  "Carp"       = "#ffd92f",
  "Gam"        = "#e5c494",
  "Gam/Tetra"  = "#b3b3b3",
  "Tetra"      = "#a6cee3",
  "Unknown"    = "#cab2d6"
)

################################################### Location + Life_S plots

# ---- Protein ----
prot_plot_LS <- compare_groups(
  data           = pr_location_cham,
  response_var   = "Protein_mg_per_g_mean",
  group_var      = "Life_S",
  subfolder_name = "Protein_LifeS"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

prot_plot_LOC <- compare_groups(
  data           = pr_location_cham,
  response_var   = "Protein_mg_per_g_mean",
  group_var      = "region",
  subfolder_name = "Protein_region"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

# ---- Phycoerythrin ----
pe_plot_LS <- compare_groups(
  data           = pe_location_cham,
  response_var   = "PE_mg_per_g_sample_mean",
  group_var      = "Life_S",
  subfolder_name = "PE_LifeS"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

pe_plot_LOC <- compare_groups(
  data           = pe_location_cham,
  response_var   = "PE_mg_per_g_sample_mean",
  group_var      = "region",
  subfolder_name = "PE_region"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

# ---- Ash ----
ash_plot_LS <- compare_groups(
  data           = ash_rep_final,
  response_var   = "ash_per_mean",
  group_var      = "Life_S",
  subfolder_name = "Ash_LifeS"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

ash_plot_LOC <- compare_groups(
  data           = ash_rep_final,
  response_var   = "ash_per_mean",
  group_var      = "region",
  subfolder_name = "Ash_region"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

# ---- Carbohydrates ----
carb_plot_LS <- compare_groups(
  data           = carb_rep_final,
  response_var   = "carb_mg_per_g_mean",
  group_var      = "Life_S",
  subfolder_name = "Carbohydrates_LifeS"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

carb_plot_LOC <- compare_groups(
  data           = carb_rep_final,
  response_var   = "carb_mg_per_g_mean",
  group_var      = "region",
  subfolder_name = "Carbohydrates_region"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

# ---- Carrageenan ----
car_plot_LS <- compare_groups(
  data           = car_rep_final,
  response_var   = "car_per",
  group_var      = "Life_S",
  subfolder_name = "Carrageenan_LifeS"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

car_plot_LOC <- compare_groups(
  data           = car_rep_final,
  response_var   = "car_per",
  group_var      = "Location",
  subfolder_name = "Carrageenan_region"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

View(car_plot_LOC)
```

## Make lists of the plots
```{r}
# Nutrients (rows)
nutrients <- c("Protein", "Carbohydrates", "ash", "Carrageenan", "Phycoerythrin")

# Two grouping variables (columns): Life_S and region
# You should have 10 plots in total
plots_lifestage <- list(prot_plot_LS, carb_plot_LS, ash_plot_LS, pe_plot_LS)
plots_location  <- list(prot_plot_LOC, carb_plot_LOC, ash_plot_LS, pe_plot_LOC )

```



#Assemble the panel
```{r}

# Unified color palette (unchanged)
group_colors <- c(
  "Mendieta"        = "#66c2a5", "Concession"      = "#66c2a5",
  "Caro Caido"      = "#fc8d62", "7H"              = "#8da0cb",
  "Powder Ilo"      = "#e78ac3", "Ilo mariculture" = "#a6d854",
  "Carp"            = "#ffd92f", "Gam"             = "#e5c494",
  "Gam/Tetra"       = "#b3b3b3", "Tetra"           = "#a6cee3",
  "Unknown"         = "#cab2d6"
)

# Display names for y-axis labeling
display_names <- c(
  "pr"   = "Protein",
  "pe"   = "Phycoerythrin",
  "carb" = "Carbohydrates",
  "ash"  = "Ash"
)

# Add y-axis labels from display_names
plots_lifestage <- purrr::imap(plots_lifestage, ~ .x + labs(y = display_names[[.y]]))
plots_location    <- purrr::imap(plots_location,    ~ .x + labs(y = display_names[[.y]]))

# --- Styling Functions ---
style_plot_general <- function(p, col) {
  p +
    scale_fill_manual(values = group_colors, guide = "none") +
    scale_color_manual(values = group_colors, guide = "none") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title         = element_blank(),
      axis.title.x       = element_blank(),
      axis.title.y       = if (col == 1) element_text(size = 18, face = "bold", margin = margin(r = 15)) else element_blank(),
      axis.text.x        = element_blank(),
      axis.text.y        = element_text(size = 14),
      axis.ticks.x       = element_blank(),
      axis.ticks.y       = element_line(size = 0.5),
      axis.line.x        = element_blank(),
      axis.line.y        = element_line(size = 0.5),
      legend.position    = "none",
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )
}

style_plot_bottom_row <- function(p) {
  p +
    theme(
      axis.title.x = element_text(size = 18, face = "bold", margin = margin(t = 15)),
      axis.text.x  = element_text(size = 14, angle = 45, hjust = 1, vjust = 1, color = "black"),
      axis.ticks.x = element_line(size = 0.5, color = "black"),
      axis.line.x  = element_line(size = 0.5, color = "black")
    )
}

# Apply general styling
plots_lifestage <- purrr::imap(plots_lifestage, ~ style_plot_general(.x, col = 1))
plots_location   <- purrr::imap(plots_location,    ~ style_plot_general(.x, col = 2))

# Safely apply bottom row styling
num_rows_ls <- length(plots_lifestage)
num_rows_rg <- length(plots_location)

if (num_rows_ls >= 1) {
  plots_lifestage[[num_rows_ls]] <- style_plot_bottom_row(plots_lifestage[[num_rows_ls]])
}
if (num_rows_rg >= 1) {
  plots_location[[num_rows_rg]] <- style_plot_bottom_row(plots_location[[num_rows_rg]])
}

# Combine plots row-wise and then vertically if lists are same length
if (num_rows_ls == num_rows_rg && num_rows_ls >= 1) {
  rows_combined <- purrr::map2(plots_lifestage, plots_location, `+`)
  full_panel <- wrap_plots(rows_combined, ncol = 1) & theme(legend.position = "none")

  ggsave(
    "output_all/plots/full_panel2.png",
    full_panel,
    width  = 14,
    height = 20,
    dpi    = 300
  )
} else {
  warning("Cannot combine plots: unequal number of plots or no plots available.")
}

# --- Legends ---
if (num_rows_ls >= 1 && num_rows_rg >= 1) {
  legend_lifstage_plot <- plots_lifestage[[num_rows_ls]] +
    scale_fill_manual(values = group_colors, name = "Life Stage") +
    scale_color_manual(values = group_colors, name = "Life Stage") +
    theme(legend.position = "right")

  legend_location_plot <- plots_location[[num_rows_rg]] +
    scale_fill_manual(values = group_colors, name = "Location") +
    scale_color_manual(values = group_colors, name = "Location") +
    theme(legend.position = "right")

  legend_lifestage <- get_legend(legend_lifstage_plot)
  legend_location    <- get_legend(legend_location_plot)

  ggsave("output_all/plots/legend_lifestage.png",
         plot = plot_grid(legend_lifestage),
         width = 4, height = 3, dpi = 300, bg = "white")

  ggsave("output_all/plots/legend_Location.png",
         plot = plot_grid(legend_location),
         width = 4, height = 3, dpi = 300, bg = "white")
} else {
  warning("Skipping legend creation due to missing plot rows.")
}
######################################
#Make the legends
######################################

num_rows_ls <- length(plots_lifestage)
num_rows_rg <- length(plots_location)

if (num_rows_ls >= 1 && num_rows_rg >= 1) {
  # Create temporary plots with legend enabled
  legend_lifestage_plot <- plots_lifestage[[num_rows_ls]] +
    scale_fill_manual(values = group_colors, name = "Group") +
    guides(fill = guide_legend(override.aes = list(color = NA))) +  # Remove outlines from legend boxes
    theme(legend.position = "right")

  legend_location_plot <- plots_location[[num_rows_rg]] +
    scale_fill_manual(values = group_colors, name = "Group") +
    guides(fill = guide_legend(override.aes = list(color = NA))) +
    theme(legend.position = "right")

  # Extract legends
  legend_lifestage <- cowplot::get_legend(legend_lifestage_plot)
  legend_location  <- cowplot::get_legend(legend_location_plot)

  # Save legends as separate images
  ggsave("output_all/plots/legend_lifestage.png",
         plot = cowplot::plot_grid(legend_lifestage),
         width = 4, height = 3, dpi = 300, bg = "white")

  ggsave("output_all/plots/legend_location.png",
         plot = cowplot::plot_grid(legend_location),
         width = 4, height = 3, dpi = 300, bg = "white")
} else {
  warning("Skipping legend creation due to missing plot rows.")
}
```
#Legend game
```{r}
library(cowplot)

# Fix legend names in the bottom row plots
legend_lifstage_plot <- plots_lifestage[[5]] +
  scale_fill_manual(values = group_colors, name = "Life_S") +
  scale_color_manual(values = group_colors, name = "Life_S") +
  theme(legend.position = "right")

legend_region_plot <- plots_region[[5]] +
  scale_fill_manual(values = group_colors, name = "region") +
  scale_color_manual(values = group_colors, name = "region") +
  theme(legend.position = "right")


#legend_lifestage <- get_legend(legend_lifestage_plot)
legend_lifestage <- get_legend(legend_lifestage_plot)
legend_region   <- get_legend(legend_region_plot)
#
# Save Life_S legend
ggsave("output_all/plots/legend_lifestage.png",
       plot = plot_grid(legend_lifestage),
       width = 4, height = 3, dpi = 300, bg = "white")

# Save region legend
ggsave("output_all/plots/legend_region.png",
       plot = plot_grid(legend_region),
       width = 4, height = 3, dpi = 300, bg = "white")


```

```{r}
# --- Merge all datasets together using common IDs ---
# We'll start with 'pe_rep_final' as a base and sequentially join the others.
# This assumes 'join_id', 'Location', 'variety', and 'Life_S' are present in all files
# and will be used as the keys for joining.

full_merged_data <- pe_rep_final %>%
  full_join(pr_rep_final, by = c("join_id"), suffix = c(".pe", ".pr")) %>%
  full_join(carb_rep_final, by = c("join_id"), suffix = c(".pr", ".carb")) %>%
  full_join(ash_rep_final, by = c("join_id"), suffix = c(".carb", ".ash")) %>%
  full_join(car_rep_final, by = c("join_id"), suffix = c(".ash", ".car"))

message("Attempting to fix character encoding in full_merged_data...")

# Identify all character columns
char_cols <- names(full_merged_data)[sapply(full_merged_data, is.character)]

# Convert each character column to UTF-8, replacing invalid sequences
# The 'sub = "byte"' option will replace invalid bytes with their byte representation,
# which can help identify them, or you can use 'sub = ""' to remove them.
for (col_name in char_cols) {
  full_merged_data[[col_name]] <- iconv(full_merged_data[[col_name]],
                                        from = "",      # Autodetect input encoding
                                        to = "UTF-8",   # Convert to UTF-8
                                        sub = "byte")   # How to handle invalid characters
                                                      # You can try 'sub = ""' to simply remove them
}
message("Character encoding fix attempt completed.")

# Check for any remaining NAs introduced by iconv (if sub = "byte" was used and characters were truly unconvertible)
# and inspect unique values of your group_vars to see if any look odd.
unique(full_merged_data$Location)
unique(full_merged_data$variety)
unique(full_merged_data$Life_S)

# Display first few rows and structure of the merged data
head(full_merged_data)

full_merged_data <- full_merged_data %>%
  mutate(
    across(
      where(is.numeric), # Selects all columns that are numeric
      ~ ifelse(is.infinite(.), NA, .) # Replaces infinite values with NA
    )
  )

write.csv(full_merged_data, "full_merged_data.csv")





```



```{r}
# List of your numeric variables for PCA/PLS-DA
# IMPORTANT: Adjust these to the actual column names in your `full_merged_data`
# based on how your merge resulted (e.g., Protein_mg_per_g_mean.PE, etc.)
numeric_vars <- c(
  "Protein_mg_per_g_mean",
  "carb_mg_per_g_mean",
  "car_per",
  "ash_per_mean",
  "PE_mg_per_g_sample_mean"
  # Add all other numeric variables from your merged data
)
group_vars <- c("Life_S", "Location", "variety")

# Returns a named vector of counts
sample_counts <- sapply(
  full_merged_data[numeric_vars],
  function(x) sum(!is.na(x))
)
print(sample_counts)

# List of your group variables
#count number of samples for different combinations
counts_by_group <- full_merged_data %>%
  group_by(across(all_of(group_vars))) %>%
  summarise(
    across(
      all_of(numeric_vars),
      ~ sum(!is.na(.)),
      .names = "n_{.col}"
    ),
    .groups = "drop"
  ) %>%
  # if you prefer a long/tidy layout:
  pivot_longer(
    cols        = starts_with("n_"),
    names_to    = "variable",
    values_to   = "n_samples",
    names_prefix= "n_"
  )

print(counts_by_group)

data_for_analysis <- full_merged_data %>%
  dplyr::select(all_of(c("join_id", group_vars, numeric_vars))) %>%
  # Combine Life_S and Location into a new variable
  mutate(LifeS_Location_Combined = paste(Life_S, Location, sep = "_")) %>%
  # Ensure the new combined variable is a factor
  mutate(LifeS_Location_Combined = as.factor(LifeS_Location_Combined)) %>%
  na.omit()

#
# BASE-R approach: a list of tables
samples_per_group <- lapply(group_vars, function(g) {
  # `useNA="no"` drops NAs; omit if you want to count NAs too
  table(full_merged_data[[g]], useNA = "no")
})
names(samples_per_group) <- group_vars

# print them out
samples_per_group
# Check the structure of the data for analysis
head(data_for_analysis)

numeric_vars_pe_pr <- c(
  "Protein_mg_per_g_mean", # Adjust if your merge gave these names
  "PE_mg_per_g_sample_mean" # Adjust if your merge gave these names
)

data_for_analysis_pe_pr <- full_merged_data %>%
  dplyr::select(all_of(c("join_id", group_vars, numeric_vars_pe_pr))) %>%
  na.omit() # Omit NA for just these two variables

message(paste("Sample size for PE/PR analysis:", nrow(data_for_analysis_pe_pr)))

```

##Biplot
```{r}
# Prepare data for PCA (numeric matrix)
pca_data <- data_for_analysis %>%
  dplyr::select(all_of(numeric_vars)) %>%
  scale() # Scale the data (important for PCA)

# Perform PCA
res.pca <- PCA(pca_data, graph = FALSE)

# Visualize scree plot (variance explained by each PC)
fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 50))

# Biplot of individuals and variables
# You can color by one of your group variables, e.g., 'Life_S'
# Biplot of individuals and variables and save
# You can color by one of your group variables, e.g., 'Life_S'
# 1. Extract numeric columns, keeping row alignment
X <- data_for_analysis[, numeric_vars]

# 2. Run PCA on just numeric data
res.pca <- prcomp(X, scale. = TRUE)

# 3. Plot, using the same data frame (rows aligned)
p_biplot <- fviz_pca_biplot(res.pca,
                            habillage = data_for_analysis$Life_S,
                            addEllipses = TRUE,
                            ellipse.type = "convex",
                            repel = TRUE,
                            ggtheme = theme_minimal())


p_biplot_combined <- fviz_pca_biplot(res.pca,
                                     habillage = data_for_analysis$LifeS_Location_Combined, # Color by combined variable
                                     addEllipses = TRUE, # Add ellipses around groups
                                     ellipse.type = "convex", # Type of ellipse
                                     repel = TRUE, # Avoid text overlapping
                                     ggtheme = theme_minimal(),
                                     title = "PCA Biplot colored by Life_S and Location") # Add a descriptive title
print(p_biplot_combined)
ggsave(filename = file.path(plot_dir, "pca_biplot_LifeS_Location_Combined.png"), plot = p_biplot_combined, width = 12, height = 10)

print(p_biplot)
ggsave(filename = file.path(plot_dir, "pca_biplot_LifeS.png"), plot = p_biplot, width = 10, height = 8)


# Contributions of variables to the first two principal components and save
p_contrib_pc1 <- fviz_contrib(res.pca, choice = "var", axes = 1, top = 10)
print(p_contrib_pc1)
ggsave(filename = file.path(plot_dir, "pca_contrib_pc1.png"), plot = p_contrib_pc1, width = 8, height = 6)

p_contrib_pc2 <- fviz_contrib(res.pca, choice = "var", axes = 2, top = 10)
print(p_contrib_pc2)
ggsave(filename = file.path(plot_dir, "pca_contrib_pc2.png"), plot = p_contrib_pc2, width = 8, height = 6)

```
##Individuals plot

```{r}
## Section 3. PCA Analysis
# ... (previous code for pca_data and res.pca) ...

# --- Individual Plot: Samples colored by Life_S ---
# This plot is excellent for visualizing clusters of samples.
p_ind_LifeS <- fviz_pca_ind(res.pca,
                            geom.ind = "point", # Show points for individuals
                            col.ind = data_for_analysis$Life_S, # Color by 'Life_S'
                            palette = "Dark2", # Choose a color palette
                            addEllipses = TRUE, # Add ellipses around groups
                            ellipse.type = "convex", # Type of ellipse
                            repel = TRUE, # Avoid text overlapping
                            legend.title = "Life_S", # Custom legend title
                            ggtheme = theme_minimal(),
                            title = "PCA Individuals Plot (Colored by Life_S)")
print(p_ind_LifeS)
ggsave(filename = file.path(plot_dir, "pca_individuals_LifeS.png"), plot = p_ind_LifeS, width = 9, height = 7)


# --- Individual Plot: Samples colored by Location ---
p_ind_Location <- fviz_pca_ind(res.pca,
                               geom.ind = "point",
                               col.ind = data_for_analysis$Location, # Color by 'Location'
                               palette = "Paired", # Another color palette example
                               addEllipses = TRUE,
                               ellipse.type = "convex",
                               repel = TRUE,
                               legend.title = "Location",
                               ggtheme = theme_minimal(),
                               title = "PCA Individuals Plot (Colored by Location)")
print(p_ind_Location)
ggsave(filename = file.path(plot_dir, "pca_individuals_Location.png"), plot = p_ind_Location, width = 9, height = 7)

# --- Individual Plot: Samples colored by the combined variable (if you created it) ---
# Assuming you added `LifeS_Location_Combined` in the previous step
if ("LifeS_Location_Combined" %in% colnames(data_for_analysis)) {
  p_ind_combined <- fviz_pca_ind(res.pca,
                                 geom.ind = "point",
                                 col.ind = data_for_analysis$LifeS_Location_Combined,
                                 addEllipses = TRUE,
                                 ellipse.type = "convex",
                                 repel = TRUE,
                                 legend.title = "Life_S and Location",
                                 ggtheme = theme_minimal(),
                                 title = "PCA Individuals Plot (Colored by Life_S & Location)")
  print(p_ind_combined)
  ggsave(filename = file.path(plot_dir, "pca_individuals_LifeS_Location_Combined.png"), plot = p_ind_combined, width = 12, height = 10)
}

# ... (rest of your PCA plots like biplot and contribution plots) ...
```
##Just protein and phycoerythrin variables

```{r}
# Prepare data for PCA
pca_data_pe_pr <- data_for_analysis_pe_pr %>%
  select(all_of(numeric_vars_pe_pr)) %>%
  scale() # Scale the data
#
group_vars <- c("Life_S", "Location", "variety")
# Perform PCA
res.pca_pe_pr <- PCA(pca_data_pe_pr, graph = FALSE)

# Scree plot
p_eig_pe_pr <- fviz_eig(res.pca_pe_pr, addlabels = TRUE, ylim = c(0, 100))
print(p_eig_pe_pr)
ggsave(filename = file.path(plot_dir, "pca_pe_pr_scree_plot.png"), plot = p_eig_pe_pr, width = 8, height = 6)

# Biplot (individuals + variables) - Colored by Life_S
p_biplot_pe_pr_LifeS <- fviz_pca_biplot(res.pca_pe_pr,
                                       habillage = data_for_analysis_pe_pr$Life_S,
                                       addEllipses = TRUE,
                                       ellipse.type = "convex",
                                       repel = TRUE,
                                       ggtheme = theme_minimal(),
                                       title = "PCA Biplot (PE/PR) - Life_S")
print(p_biplot_pe_pr_LifeS)
ggsave(filename = file.path(plot_dir, "pca_pe_pr_biplot_LifeS.png"), plot = p_biplot_pe_pr_LifeS, width = 10, height = 8)

# Biplot (individuals + variables) - Colored by Location
p_biplot_pe_pr_Location <- fviz_pca_biplot(res.pca_pe_pr,
                                          habillage = data_for_analysis_pe_pr$Location,
                                          addEllipses = TRUE,
                                          ellipse.type = "convex",
                                          repel = TRUE,
                                          ggtheme = theme_minimal(),
                                          title = "PCA Biplot (PE/PR) - Location")
print(p_biplot_pe_pr_Location)
ggsave(filename = file.path(plot_dir, "pca_pe_pr_biplot_Location.png"), plot = p_biplot_pe_pr_Location, width = 10, height = 8)


# Individual Plot (just samples) - Colored by Life_S
p_ind_pe_pr_LifeS <- fviz_pca_ind(res.pca_pe_pr,
                                  geom.ind = "point",
                                  col.ind = data_for_analysis_pe_pr$Life_S,
                                  addEllipses = TRUE,
                                  ellipse.type = "convex",
                                  repel = TRUE,
                                  ggtheme = theme_minimal(),
                                  title = "PCA Individuals (PE/PR) - Life_S")
print(p_ind_pe_pr_LifeS)
ggsave(filename = file.path(plot_dir, "pca_pe_pr_individuals_LifeS.png"), plot = p_ind_pe_pr_LifeS, width = 9, height = 7)

# Individual Plot (just samples) - Colored by Location
p_ind_pe_pr_Location <- fviz_pca_ind(res.pca_pe_pr,
                                   geom.ind = "point",
                                   col.ind = data_for_analysis_pe_pr$Location,
                                   addEllipses = TRUE,
                                   ellipse.type = "convex",
                                   repel = TRUE,
                                   ggtheme = theme_minimal(),
                                   title = "PCA Individuals (PE/PR) - Location")
print(p_ind_pe_pr_Location)
ggsave(filename = file.path(plot_dir, "pca_pe_pr_individuals_Location.png"), plot = p_ind_pe_pr_Location, width = 9, height = 7)
```
##Summary tables 


```{r beautiful_summary_all_nutrients_tables_colored}
library(dplyr)
library(tidyr)
library(gt)
library(rlang)

# Define numeric variables for summary
numeric_vars_all_nutrients <- c(
  "Protein_mg_per_g_mean",
  "PE_mg_per_g_sample_mean",
  "carb_mg_per_g_mean",
  "car_per",
  "ash_per_mean"
)

# Pretty names for display
display_names <- c(
  "Protein_mg_per_g_mean" = "Protein (mg/g)",
  "PE_mg_per_g_sample_mean" = "Phycoerythrin (mg/g)",
  "carb_mg_per_g_mean" = "Carbohydrates (mg/g)",
  "car_per" = "Carrageenan (%)",
  "ash_per_mean" = "Ash (%)"
)

# Read your full dataset
full_merged_data <- read.csv("input_all/full_merged_data_manual_fix.csv")

# Clean and summarize function
summarize_clean_numeric <- function(df, group_var, numeric_vars) {
  df_clean <- df
  for (var in numeric_vars) {
    x <- as.numeric(df_clean[[var]])
    x[is.nan(x) | is.infinite(x)] <- NA
    x[is.na(x)] <- mean(x, na.rm = TRUE)
    df_clean[[var]] <- x
  }

  summary_df <- df %>%
    dplyr::select(all_of(c(group_var, numeric_vars))) %>%
    tidyr::pivot_longer(cols = all_of(numeric_vars), names_to = "Variable", values_to = "Value") %>%
    dplyr::group_by(.data[[group_var]], Variable) %>%
    dplyr::summarise(
      Mean = mean(Value, na.rm = TRUE),
      SE = sd(Value, na.rm = TRUE) / sqrt(length(na.omit(Value))),
      N = length(na.omit(Value)),
      .groups = "drop"
    )

  for (col in c("Mean", "SE")) {
    x <- summary_df[[col]]
    x[is.nan(x) | is.infinite(x)] <- NA
    x[is.na(x)] <- mean(x, na.rm = TRUE)
    summary_df[[col]] <- x
  }

  group_sym <- sym(group_var)
  summary_df <- summary_df %>%
    filter(!is.na(!!group_sym) & !is.nan(!!group_sym) & !!group_sym != "")

  # Ensure all strings are valid UTF-8
  summary_df <- summary_df %>%
    mutate(across(where(is.character), ~ iconv(., from = "", to = "UTF-8", sub = "byte")))

  return(summary_df)
}

# Clean and summarize by group
summary_LifeS_long     <- summarize_clean_numeric(full_merged_data, "Life_S", numeric_vars_all_nutrients)
summary_Location_long  <- summarize_clean_numeric(full_merged_data, "Location", numeric_vars_all_nutrients)
summary_Variety_long   <- summarize_clean_numeric(full_merged_data, "variety", numeric_vars_all_nutrients)

# Define global color scale
all_means_for_coloring_all_nutrients <- c(
  summary_LifeS_long$Mean,
  summary_Location_long$Mean,
  summary_Variety_long$Mean
)
global_min_mean_all_nutrients <- min(all_means_for_coloring_all_nutrients, na.rm = TRUE)
global_max_mean_all_nutrients <- max(all_means_for_coloring_all_nutrients, na.rm = TRUE)

# gt() table generator function
make_gt_table <- function(summary_df, group_var_label, title_label, colorize = TRUE) {
  gt_table <- summary_df %>%
    pivot_wider(names_from = Variable, values_from = c(Mean, SE, N)) %>%
    gt()

  if (colorize) {
    gt_table <- gt_table %>%
      data_color(
        columns = starts_with("Mean_"),
        palette = "Blues",
        domain = c(global_min_mean_all_nutrients, global_max_mean_all_nutrients)
      )
  }

  gt_table <- gt_table %>%
    {
      gt_obj <- .
      for (var_name in numeric_vars_all_nutrients) {
        mean_col <- sym(paste0("Mean_", var_name))
        se_col <- sym(paste0("SE_", var_name))
        n_col  <- sym(paste0("N_", var_name))
        if (all(as.character(c(mean_col, se_col, n_col)) %in% names(gt_obj[["_data"]]))) {
          gt_obj <- gt_obj %>%
            fmt_number(columns = c(!!mean_col, !!se_col), decimals = 2) %>%
            fmt_integer(columns = !!n_col) %>%
            cols_merge(columns = c(!!mean_col, !!se_col, !!n_col),
                       pattern = "{1} ± {2} (N={3})") %>%
            cols_label(!!mean_col := display_names[[var_name]])
        }
      }
      gt_obj
    } %>%
    cols_label(!!sym(group_var_label) := md(paste0("**", gsub("_", " ", group_var_label), "**"))) %>%
    tab_header(title = md(paste0("**Summary of All Nutrients by ", title_label, "**"))) %>%
    opt_horizontal_padding(scale = 2) %>%
    tab_options(
      table.border.top.color = "lightgray",
      table.border.bottom.color = "lightgray",
      heading.border.bottom.color = "lightgray",
      column_labels.border.bottom.color = "lightgray"
    ) %>%
    tab_footnote(
      footnote = "Values represent Mean ± Standard Error (SE) (N = number of samples).",
      locations = cells_column_labels(columns = starts_with("Mean_"))
    )

  return(gt_table)
}

# Generate gt tables
gt_table_LifeS_colored    <- make_gt_table(summary_LifeS_long, "Life_S", "Life_S", colorize=FALSE)
gt_table_Location_colored <- make_gt_table(summary_Location_long, "Location", "Location", colorize=FALSE)
gt_table_Variety_colored  <- make_gt_table(summary_Variety_long, "variety", "Variety", colorize=FALSE)

clean_gt_data_utf8 <- function(gt_obj) {
  gt_obj$`_data` <- gt_obj$`_data` %>%
    mutate(across(where(is.character), ~ iconv(., from = "", to = "UTF-8", sub = "byte")))
  gt_obj
}

# Apply it to all gt tables
gt_table_LifeS_colored    <- clean_gt_data_utf8(gt_table_LifeS_colored)
gt_table_Location_colored <- clean_gt_data_utf8(gt_table_Location_colored)
gt_table_Variety_colored  <- clean_gt_data_utf8(gt_table_Variety_colored)

gt_table_LifeS_colored
# Save tables
gtsave(gt_table_LifeS_colored, file.path(data_dir, "summary_all_nutrients_LifeS_table_colored.html"))
gtsave(gt_table_LifeS_colored, file.path(plot_dir, "summary_all_nutrients_LifeS_table_colored.png"))

gtsave(gt_table_Location_colored, file.path(data_dir, "summary_all_nutrients_Location_table_colored.html"))
gtsave(gt_table_Location_colored, file.path(plot_dir, "summary_all_nutrients_Location_table_colored.png"))

gtsave(gt_table_Variety_colored, file.path(data_dir, "summary_all_nutrients_Variety_table_colored.html"))
gtsave(gt_table_Variety_colored, file.path(plot_dir, "summary_all_nutrients_Variety_table_colored.png"))


```


