author: "Trevor Eakes"
date: "`r Sys.Date()`"
title: "PCA and PCR proximate nutrient analysis and graphing"
output: html_document
params:
  plot_dir:    "output_all/plots"
  data_dir:    "output_all/export_data"
  report_dir:  "output_all/reports"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load script and directories
```{r}
# Source script from top-level project folder
source("../process_plate_run.R")

#— 1. pull from params (or use defaults)
plot_dir   <-  "output_all/plots"
data_dir   <-  "output_all/export_data"
report_dir <-  "output_all/reports"

#— 2. create all dirs
walk(
  c(plot = plot_dir, data = data_dir, reports = report_dir),
  ~ if (!dir_exists(.x)) dir_create(.x, recurse = TRUE)
)

#— 3. register plot_dir with knitr
opts_chunk$set(fig.path = paste0(plot_dir, "/"))
#
```

## Load input data
```{r load-excel-files, warning=FALSE}
# 1. Define the folder containing all input files
input_folder <- "input_all"

# 2. List all Excel and CSV files (full paths), excluding temp Excel files (~$)
all_files <- list.files(
  path = input_folder,
  pattern = "\\.(xlsx|csv)$",
  full.names = TRUE
)
all_files <- all_files[!grepl("^~\\$", basename(all_files))]

# 3. Loop over each file
for (file in all_files) {
  # 3a. Create a clean object name (strip out folder & extension)
  name <- file_path_sans_ext(basename(file))

  # 3b. Determine file type
  if (grepl("\\.xlsx$", file)) {
    # Read Excel file (2nd sheet if available, else 1st)
    sheet_names   <- excel_sheets(file)
    sheet_to_read <- if (length(sheet_names) >= 2) sheet_names[2] else sheet_names[1]
    data <- suppressMessages(read_excel(file, sheet = sheet_to_read))
    message("Loaded Excel file: ", name, " (sheet = '", sheet_to_read, "')")
  } else if (grepl("\\.csv$", file)) {
    # Read CSV file
    data <- suppressMessages(read_csv(file, show_col_types = FALSE))
    message("Loaded CSV file: ", name)
  }

  # 3c. Assign the data.frame to the global environment under "name"
  assign(name, data, envir = .GlobalEnv)
}
```
## Make the factors and color levels
```{r}
# For Life Stage
# --- Define your variables
group_vars <- c("Life_S", "Location", "variety")
numeric_vars <- c(
  "Protein_mg_per_g_mean", "PE_mg_per_g_sample_mean", "carb_mg_per_g_mean",
  "ash_per_mean", "car_per"
)

# --- Load and clean data
data <- readxl::read_excel("full_merged_data_manual_fix.xlsx") %>%
  mutate(
    Lab_Num = as.numeric(str_extract(join_id, "\\d+")),
    across(all_of(numeric_vars), ~ as.numeric(.))
  ) %>%
  dplyr::filter(str_detect(join_id, "^Lab_")) #%>%
  #dplyr::filter(Lab_Num >= 1 & Lab_Num <= 22)

#Code to collapse factors
desired_location_order <- c("Mendieta", "Concession", "Caro Caido", "7H", "Powder Ilo", "Ilo mariculture")
life_s_levels           <- c("Carp", "Gam", "Gam/Tetra", "Tetra", "Unknown")
 # or whatever order you want

```

## Make the plots
```{r}

# Apply color palette
group_colors <- c(
  # Location colors
  "Mendieta"        = "#66c2a5",
  "Concession"      = "#66c2a5",
  "Carro Caido"      = "#fc8d62",
  "7H"              = "#8da0cb",
  "Powder Ilo"      = "#e78ac3",
  "Ilo mariculture" = "#a6d854",
  # Life Stage colors
  "Carp"       = "#ffd92f",
  "Gam"        = "#e5c494",
  "Gam/Tetra"  = "#b3b3b3",
  "Tetra"      = "#a6cee3",
  "Unknown"    = "#cab2d6",
  #Region colors
    "Paracas"        = "#8da0cb",
  "Marcona"             = "#e78ac3",
  "Ilo" = "#a6d854"
  
)

################################################### Location + Life Stage plots

# ---- Protein ----
prot_plot_LS <- compare_groups(
  data           = data,
  response_var   = "Protein_mg_per_g_mean",
  group_var      = "Life_S",
  subfolder_name = "Protein_LifeS"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

prot_plot_LOC <- compare_groups(
  data           = data,
  response_var   = "Protein_mg_per_g_mean",
  group_var      = "Location",
  subfolder_name = "Protein_region"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

# ---- Phycoerythrin ----
pe_plot_LS <- compare_groups(
  data           = data,
  response_var   = "PE_mg_per_g_sample_mean",
  group_var      = "Life_S",
  subfolder_name = "PE_LifeS"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

pe_plot_LOC <- compare_groups(
  data           = data,
  response_var   = "PE_mg_per_g_sample_mean",
  group_var      = "Location",
  subfolder_name = "PE_location"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

# ---- Ash ----
ash_plot_LS <- compare_groups(
  data           = data,
  response_var   = "ash_per_mean",
  group_var      = "Life_S",
  subfolder_name = "Ash_LifeS"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

ash_plot_LOC <- compare_groups(
  data           = data,
  response_var   = "ash_per_mean",
  group_var      = "region",
  subfolder_name = "Ash_region"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

# ---- Carbohydrates ----
carb_plot_LS <- compare_groups(
  data           = data,
  response_var   = "carb_mg_per_g_mean",
  group_var      = "Life_S",
  subfolder_name = "Carbohydrates_LifeS"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

carb_plot_LOC <- compare_groups(
  data           = data,
  response_var   = "carb_mg_per_g_mean",
  group_var      = "region",
  subfolder_name = "Carbohydrates_region"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

# ---- Carrageenan ----
car_plot_LS <- compare_groups(
  data           = data,
  response_var   = "car_per",
  group_var      = "Life_S",
  subfolder_name = "Carrageenan_LifeS"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

car_plot_LOC <- compare_groups(
  data           = data,
  response_var   = "car_per",
  group_var      = "region",
  subfolder_name = "Carrageenan_region"
) + scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors)

```

## Make lists of the plots
```{r}
# Nutrients (rows)
display_names <- c(
  "Protein_mg_per_g_mean" = "Protein (mg/g)",
  "carb_mg_per_g_mean" = "Carbohydrates (mg/g)",
  "ash_per_mean" = "Ash (%)",
  "PE_mg_per_g_sample_mean" = "Phycoerythrin (mg/g)",
  "car_per" = "Carrageenan (%)"
)

# Two grouping variables (columns): Life_Stage and region
# You should have 10 plots in total
plots_lifestage <- list(prot_plot_LS, carb_plot_LS, ash_plot_LS, pe_plot_LS, car_plot_LS)
plots_location  <- list(prot_plot_LOC, carb_plot_LOC, ash_plot_LOC, pe_plot_LOC, car_plot_LOC )

```



#Assemble the panel
```{r}

# 1) Your NEW unified vivid color palette (unchanged)
group_colors <- c(
  "Mendieta"        = "#FF6347", # Tomato (vivid red)
  "Concession"      = "#4682B4", # SteelBlue (strong blue)
  "Carro Caido"     = "#3CB371", # MediumSeaGreen (bright green)
  "7H"              = "#9932CC", # DarkOrchid (deep purple)
  "Powder Ilo"      = "#FFD700", # Gold (vivid yellow)
  "Ilo mariculture" = "#00CED1", # DarkTurquoise (bright cyan)
  "Carp"            = "#FF4500", # OrangeRed (intense orange)
  "Gam"             = "#1E90FF", # DodgerBlue (another strong blue)
  "Gam/Tetra"       = "#FF69B4", # HotPink (vivid pink)
  "Tetra"           = "#ADFF2F", # GreenYellow (bright lime green)
  "Paracas"         = "#808080",  # Grey (kept neutral for 'Unknown')
  "Marcona"         = "#8da0cb",
  "Ilo"             = "#e78ac3"
)

# Helper function to insert a line break after first space or slash
wrap_label <- function(x) {
  sub("([/ ]+)", "\n", x)  # Replace first space or slash with newline
}

# Function to apply styling (now only handles appearance/visibility of titles, not their text)
style_plot <- function(p, row, col, total_rows) {
  p +
    # Wrap x-axis labels (for grouped plots)
    scale_x_discrete(labels = wrap_label) +

    # Colors
    scale_fill_manual(values = group_colors, guide = "none") +

    theme_minimal(base_size = 14) +
    theme(
      plot.title         = element_blank(),

      # X-axis title visibility
      axis.title.x       = if (row == total_rows) {
                             element_text(size = 18, face = "bold", margin = margin(t = 15))
                           } else {
                             element_blank()
                           },

      # Y-axis title only for left column
      axis.title.y       = if (col == 1) element_text(size = 18, face = "bold", margin = margin(r = 15)) else element_blank(),

      # Tick labels
      axis.text.x        = element_text(size = 14, angle = 45, hjust = 0.5, vjust = 0.5, color = "black"),
      axis.text.y        = element_text(size = 14),
      axis.ticks.x       = element_line(size = 0.5, color = "black"),
      axis.ticks.y       = element_line(size = 0.5),
      axis.line.x        = element_line(size = 0.5, color = "black"),
      axis.line.y        = element_line(size = 0.5),
      legend.position    = "none",
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )
}

# Determine the total number of rows (assuming plots_lifestage holds all rows for one column)
num_rows <- length(plots_lifestage)

# 3) Apply styling to all plots, passing row and column indices
# This applies the general theme and makes x-axis title *visible* on the bottom row,
# but doesn't set the *text* yet.
plots_lifestage <- purrr::imap(plots_lifestage,
  ~ style_plot(.x, row = .y, col = 1, total_rows = num_rows) # Pass row index (.y) and col = 1
)
# Apply display_names to the y-axis labels of the left column (lifestage plots)
nutrient_names <- names(display_names)

plots_lifestage <- purrr::imap(plots_lifestage, function(p, i) {
  p + labs(y = display_names[[nutrient_names[i]]])
})
plots_location <- purrr::imap(plots_location,
  ~ style_plot(.x, row = .y, col = 2, total_rows = num_rows) # Pass row index (.y) and col = 2
)

# 4) Explicitly set the x-axis title text for ONLY the bottom plots
# This step is crucial for setting the actual text.
plots_lifestage[[num_rows]] <- plots_lifestage[[num_rows]] + labs(x = "Life Stage")
plots_location[[num_rows]] <- plots_location[[num_rows]] + labs(x = "Location")

# 5) Combine the plots horizontally for each row
rows_combined <- purrr::map2(plots_lifestage, plots_location, `+`)

# 6) Stack the combined rows vertically to form the full panel
full_panel <- wrap_plots(rows_combined, ncol = 1)

# 7) Final check: Ensure no legends are present in the combined panel
full_panel <- full_panel & theme(legend.position = "none")

# 8) Save high-res
ggsave(
  "output_all/plots/full_panel.png",
  full_panel,
  width  = 14,    # inches
  height = 20,
  dpi    = 300
)
```


#Make the big tables
```{r beautiful_summary_all_nutrients_tables_colored}
library(dplyr)
library(tidyr)
library(gt)
library(rlang)

# Define numeric variables for summary
numeric_vars_all_nutrients <- c(
  "Protein_mg_per_g_mean",
  "PE_mg_per_g_sample_mean",
  "carb_mg_per_g_mean",
  "car_per",
  "ash_per_mean"
)

# Pretty names for display
display_names <- c(
  "Protein_mg_per_g_mean" = "Protein (mg/g)",
  "PE_mg_per_g_sample_mean" = "Phycoerythrin (mg/g)",
  "carb_mg_per_g_mean" = "Carbohydrates (mg/g)",
  "car_per" = "Carrageenan (%)",
  "ash_per_mean" = "Ash (%)"
)

# Read your full dataset
full_merged_data <- read_xlsx("full_merged_data_manual_fix.xlsx") %>%
  mutate(across(all_of(numeric_vars), ~ as.numeric(gsub(",", ".", gsub("[^0-9.,-]", "", .)))))

# Clean and summarize function
summarize_clean_numeric <- function(df, group_var, numeric_vars) {
  df_clean <- df
  for (var in numeric_vars) {
    x <- as.numeric(df_clean[[var]])
    x[is.nan(x) | is.infinite(x)] <- NA
    x[is.na(x)] <- mean(x, na.rm = TRUE)
    df_clean[[var]] <- x
  }

  summary_df <- df %>%
    dplyr::select(all_of(c(group_var, numeric_vars))) %>%
    tidyr::pivot_longer(cols = all_of(numeric_vars), names_to = "Variable", values_to = "Value") %>%
    dplyr::group_by(.data[[group_var]], Variable) %>%
    dplyr::summarise(
      Mean = mean(Value, na.rm = TRUE),
      SE = sd(Value, na.rm = TRUE) / sqrt(length(na.omit(Value))),
      N = length(na.omit(Value)),
      .groups = "drop"
    )

  for (col in c("Mean", "SE")) {
    x <- summary_df[[col]]
    x[is.nan(x) | is.infinite(x)] <- NA
    x[is.na(x)] <- mean(x, na.rm = TRUE)
    summary_df[[col]] <- x
  }

  group_sym <- sym(group_var)
  summary_df <- summary_df %>%
    dplyr::filter(!is.na(!!group_sym) & !is.nan(!!group_sym) & !!group_sym != "")

  # Ensure all strings are valid UTF-8
  summary_df <- summary_df %>%
    mutate(across(where(is.character), ~ iconv(., from = "", to = "UTF-8", sub = "byte")))

  return(summary_df)
}

# Clean and summarize by group
summary_LifeS_long     <- summarize_clean_numeric(full_merged_data, "Life_S", numeric_vars_all_nutrients)
summary_Location_long  <- summarize_clean_numeric(full_merged_data, "Location", numeric_vars_all_nutrients)
summary_Variety_long   <- summarize_clean_numeric(full_merged_data, "variety", numeric_vars_all_nutrients)

# Define global color scale
all_means_for_coloring_all_nutrients <- c(
  summary_LifeS_long$Mean,
  summary_Location_long$Mean,
  summary_Variety_long$Mean
)
global_min_mean_all_nutrients <- min(all_means_for_coloring_all_nutrients, na.rm = TRUE)
global_max_mean_all_nutrients <- max(all_means_for_coloring_all_nutrients, na.rm = TRUE)

# gt() table generator function
make_gt_table <- function(summary_df, group_var_label, title_label, colorize = TRUE) {
  gt_table <- summary_df %>%
    pivot_wider(names_from = Variable, values_from = c(Mean, SE, N)) %>%
    gt()

  if (colorize) {
    gt_table <- gt_table %>%
      data_color(
        columns = starts_with("Mean_"),
        palette = "Blues",
        domain = c(global_min_mean_all_nutrients, global_max_mean_all_nutrients)
      )
  }

  gt_table <- gt_table %>%
    {
      gt_obj <- .
      for (var_name in numeric_vars_all_nutrients) {
        mean_col <- sym(paste0("Mean_", var_name))
        se_col <- sym(paste0("SE_", var_name))
        n_col  <- sym(paste0("N_", var_name))
        if (all(as.character(c(mean_col, se_col, n_col)) %in% names(gt_obj[["_data"]]))) {
          gt_obj <- gt_obj %>%
            fmt_number(columns = c(!!mean_col, !!se_col), decimals = 2) %>%
            fmt_integer(columns = !!n_col) %>%
            cols_merge(columns = c(!!mean_col, !!se_col, !!n_col),
                       pattern = "{1} ± {2} (N={3})") %>%
            cols_label(!!mean_col := display_names[[var_name]])
        }
      }
      gt_obj
    } %>%
    cols_label(!!sym(group_var_label) := md(paste0("**", gsub("_", " ", group_var_label), "**"))) %>%
    tab_header(title = md(paste0("**Summary of All Nutrients by ", title_label, "**"))) %>%
    opt_horizontal_padding(scale = 2) %>%
    tab_options(
      table.border.top.color = "lightgray",
      table.border.bottom.color = "lightgray",
      heading.border.bottom.color = "lightgray",
      column_labels.border.bottom.color = "lightgray"
    ) %>%
    tab_footnote(
      footnote = "Values represent Mean ± Standard Error (SE) (N = number of samples).",
      locations = cells_column_labels(columns = starts_with("Mean_"))
    )

  return(gt_table)
}

# Generate gt tables
gt_table_LifeS_colored    <- make_gt_table(summary_LifeS_long, "Life_S", "Life_S", colorize=FALSE)
gt_table_Location_colored <- make_gt_table(summary_Location_long, "Location", "Location", colorize=FALSE)
gt_table_Variety_colored  <- make_gt_table(summary_Variety_long, "variety", "Variety", colorize=FALSE)

clean_gt_data_utf8 <- function(gt_obj) {
  gt_obj$`_data` <- gt_obj$`_data` %>%
    mutate(across(where(is.character), ~ iconv(., from = "", to = "UTF-8", sub = "byte")))
  gt_obj
}

# Apply it to all gt tables
gt_table_LifeS_colored    <- clean_gt_data_utf8(gt_table_LifeS_colored)
gt_table_Location_colored <- clean_gt_data_utf8(gt_table_Location_colored)
gt_table_Variety_colored  <- clean_gt_data_utf8(gt_table_Variety_colored)

gt_table_LifeS_colored
# Save tables
gtsave(gt_table_LifeS_colored, file.path(data_dir, "summary_all_nutrients_LifeS_table_colored.html"))
gtsave(gt_table_LifeS_colored, file.path(plot_dir, "summary_all_nutrients_LifeS_table_colored.png"))

gtsave(gt_table_Location_colored, file.path(data_dir, "summary_all_nutrients_Location_table_colored.html"))
gtsave(gt_table_Location_colored, file.path(plot_dir, "summary_all_nutrients_Location_table_colored.png"))

gtsave(gt_table_Variety_colored, file.path(data_dir, "summary_all_nutrients_Variety_table_colored.html"))
gtsave(gt_table_Variety_colored, file.path(plot_dir, "summary_all_nutrients_Variety_table_colored.png"))


```