---
title: "Carrageenan classification"
author: "Trevor Eakes"
date: "`r Sys.Date()`"
output: html_document
---

Explanation of key steps

Max‐peak extraction (df_peaks): for each sample and each band range, we pull the single highest absorbance.

Arithmetic‐means normalization (df_norm): each band’s peak is divided by the mean of that sample’s peaks, yielding relative intensities around 1.

Index calculation (scores):

Kappa = G4S / DA

Iota = DA2S / DA

Lambda = (G2S + DA2S) / DA

Final tables:

Table3_Intensities: one row per band, showing relative intensities for each sample.

Carrageenan_Scores: per‐sample continuous Kappa, Iota, and Lambda indices.
```{r}
# Required packages


# 1) Read your updated data
df_raw <- read_csv("metaboanalyst_main_input_data.csv")

# 2) Convert to long format (only spectral columns)
df_ftir <- df_raw %>%
  pivot_longer(
    cols = matches("^[0-9]{3,4}$"),  # selects columns like "4000", "3999", etc.
    names_to = "Wavenumber",
    values_to = "Absorbance"
  ) %>%
  mutate(Wavenumber = as.numeric(Wavenumber))

# 2) Define your carrageenan bands
bands <- tribble(
  ~code,     ~name,                                   ~lo,   ~hi,
  "S",       "Sulphate ester (O–SO₃⁻)",              1205,  1265,
  "DA1070",  "C–O band of 3,6-anhydrogalactose",     1055,  1075,
  "DA930",   "3,6-anhydro-D-galactose",              920,   935,
  "G4S",     "D-galactose-4-sulphate",               840,   850,
  "G2S",     "D-galactose-2-sulphate",               825,   830,
  "D2S",     "D-galactose-2,6-disulphate",           815,   820,
  "DA2S",    "3,6-anhydro-D-galactose-2-sulphate",   800,   805
)

# 3) Extract max-peak intensity for each SAMPLE+GROUP & band
df_peaks <- df_ftir %>%
  crossing(bands) %>%
  filter(Wavenumber >= lo, Wavenumber <= hi) %>%
  group_by(Sample, code, name, lo, hi) %>%
  slice_max(order_by = Absorbance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(Peak = Absorbance, Peak_Position = Wavenumber)

df_peaks %>%
  group_by(Sample, code) %>%
  summarize(n = n(), .groups = "drop") %>%
  filter(n > 1)

#
```

```{r}
# --- Filter only the spectral range of interest (1400 to 700) ---
# Create output folder if needed
plot_dir <- "output_ftir/plots"
dir_create(plot_dir, recurse = TRUE)

# Filter bands to only those within 1400–700
bands_subset <- bands %>% filter(lo <= 1400, hi >= 700)

# Filter FTIR data to the same range
df_ftir_subset <- df_ftir %>%
  filter(Wavenumber <= 1400, Wavenumber >= 700)

# Plot function for one sample
plot_individual_spectrum <- function(sample_id) {
  df_sample <- df_ftir_subset %>% filter(Sample == sample_id)

  p <- ggplot(df_sample, aes(x = Wavenumber, y = Absorbance)) +
    geom_line(color = "black", linewidth = 1) +
    scale_x_reverse() +
    labs(
      title = paste("FTIR Spectrum:", sample_id),
      x = "Wavenumber (cm⁻¹)",
      y = "Absorbance"
    ) +
    theme_classic(base_size = 13)

  for (i in seq_len(nrow(bands_subset))) {
    band <- bands_subset[i, ]
    center <- (band$lo + band$hi) / 2
    band_height <- max(df_sample$Absorbance, na.rm = TRUE)

    p <- p +
      annotate("rect", xmin = band$lo, xmax = band$hi,
               ymin = -Inf, ymax = Inf, alpha = 0.15, fill = "steelblue") +
      annotate("text", x = center, y = band_height * 0.95,
               label = band$code, angle = 90, vjust = -0.5, size = 3.5)
  }

  ggsave(
    filename = file.path(plot_dir, paste0("spectrum_", sample_id, ".png")),
    plot = p, width = 8, height = 5, bg = "white"
  )
}

# Apply to all samples
walk(unique(df_ftir_subset$Sample), plot_individual_spectrum)

message("✅ Individual sample plots saved with shaded bands.")

```


```{r}
# 4) Pivot so each 'code' is its own column
df_peak_wide <- df_peaks %>%
  select(Sample, code, Peak) %>%
  pivot_wider(
    names_from  = code,
    values_from = Peak
  ) %>%
  # fill any missing (NA) peaks with 0
  mutate(across(-c(Sample), ~ replace_na(.x, 0)))

# 5) Arithmetic-mean normalization per sample
val_cols <- df_peak_wide %>%
  select(where(is.numeric)) %>%
  names()
df_norm <- df_peak_wide %>%
  rowwise() %>%
  mutate(mean_peak = mean(c_across(all_of(val_cols)), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(across(
    all_of(val_cols),
    ~ ifelse(mean_peak > 0, .x / mean_peak, 0)
  )) %>%
  select(-mean_peak)

# 6) Calculate carrageenan indices and merge everything

# Define the metadata columns you want to carry over
meta_cols <- c("Sample", "Sample Code", "Location", "region", "variety", "Life_S", "car_per")

# Extract unique metadata from the original input
metadata_only <- df_raw %>%
  select(all_of(meta_cols)) %>%
  distinct()

# Join metadata to normalized spectral data and calculate indices
final_table <- df_norm %>%
  left_join(metadata_only, by = "Sample") %>%
  relocate(all_of(meta_cols)) %>%
  mutate(
    Kappa_Index  = if_else(DA1070 > 0, G4S / DA1070, NA_real_),
    Iota_Index   = if_else(DA1070 > 0, DA2S / DA1070, NA_real_),
    Lambda_Index = if_else(DA1070 > 0, (G2S + DA2S) / DA1070, NA_real_)
  ) %>%
  select(all_of(meta_cols), all_of(val_cols), Kappa_Index, Iota_Index, Lambda_Index)

# 7) Write one combined output table
write_csv(final_table, "ftir_combined_results.csv")

# 8) Print preview
cat("\n=== Combined Table: Metadata + Intensities + Scores ===\n")
print(final_table)


# 9) Print to console for quick check
cat("\n=== Table 3 — Normalized Band Intensities ===\n")
print(table3)

cat("\n=== Carrageenan Relative Scores ===\n")
print(scores)

library(tidyverse)

# --- 1. Read data ---
df_raw <- read_csv("metaboanalyst_main_input_data.csv")

# --- 2. Define metadata columns explicitly ---
meta_cols <- c("Sample", "Sample Code", "Location", "region", "variety", "Life_S", "car_per")

# --- 3. Convert to long format for FTIR bands ---
df_ftir <- df_raw %>%
  rename(SAMPLE = Sample) %>%  # Sample Code used as GROUP here
  pivot_longer(
    cols = matches("^[0-9]{3,4}$"),  # match numeric wavenumber columns like "4000"
    names_to = "Wavenumber",
    values_to = "Absorbance"
  ) %>%
  mutate(Wavenumber = as.numeric(Wavenumber))

# --- 4. Define carrageenan bands ---
bands <- tribble(
  ~code,     ~name,                                   ~lo,   ~hi,
  "S",       "Sulphate ester (O–SO₃⁻)",              1210,  1260,
  "DA1070",  "C–O band of 3,6-anhydrogalactose",     1070,  1070,
  "DA930",   "3,6-anhydro-D-galactose",              928,   933,
  "G4S",     "D-galactose-4-sulphate",               840,   850,
  "G2S",     "D-galactose-2-sulphate",               825,   830,
  "D2S",     "D-galactose-2,6-disulphate",           815,   820,
  "DA2S",    "3,6-anhydro-D-galactose-2-sulphate",   800,   805
)

# --- 5. Extract max peak intensity per band per sample ---
df_peaks <- df_ftir %>%
  crossing(bands) %>%
  filter(Wavenumber >= lo, Wavenumber <= hi) %>%
  group_by(SAMPLE, GROUP, code, name, lo, hi) %>%
  slice_max(order_by = Absorbance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(Peak = Absorbance, Peak_Position = Wavenumber)

# --- 6. Pivot wide: each band as a column ---
df_peak_wide <- df_peaks %>%
  select(SAMPLE, GROUP, code, Peak) %>%
  pivot_wider(
    names_from  = code,
    values_from = Peak
  ) %>%
  mutate(across(-c(SAMPLE, GROUP), ~ replace_na(.x, 0)))  # fill missing with 0

# --- 7. Normalize by mean peak intensity ---
val_cols <- setdiff(names(df_peak_wide), c("SAMPLE", "GROUP"))
df_norm <- df_peak_wide %>%
  rowwise() %>%
  mutate(mean_peak = mean(c_across(all_of(val_cols)), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(across(
    all_of(val_cols),
    ~ ifelse(mean_peak > 0, .x / mean_peak, 0)
  )) %>%
  select(-mean_peak)

# --- 8. Merge metadata from original file ---
metadata_only <- df_raw %>%
  select(all_of(meta_cols)) %>%
  distinct()

final_table <- df_norm %>%
  left_join(metadata_only, by = c("SAMPLE" = "Sample")) %>%
  relocate(all_of(meta_cols)) %>%
  mutate(
    Kappa_Index  = if_else(DA1070 > 0, G4S / DA1070, NA_real_),
    Iota_Index   = if_else(DA1070 > 0, DA2S / DA1070, NA_real_),
    Lambda_Index = if_else(DA1070 > 0, (G2S + DA2S) / DA1070, NA_real_)
  )

# --- 9. Write output ---
write_csv(final_table, "ftir_combined_results.csv")

# --- 10. Print preview ---
cat("\n=== Combined Table: Metadata + Intensities + Scores ===\n")
print(final_table %>% select(all_of(meta_cols), Kappa_Index, Iota_Index, Lambda_Index))

cat("\n=== Table 3 — Normalized Band Intensities ===\n")
print(final_table %>% select(SAMPLE, all_of(val_cols)))

cat("\n=== Carrageenan Relative Scores ===\n")
print(final_table %>% select(SAMPLE, Kappa_Index, Iota_Index, Lambda_Index))

```
```{r}

# Step 2: call mixOmics from Bioconductor
library("mixOmics")

# --- Define input and output ---
data_path <- "ftir_combined_results_cleaned.csv"
plot_dir  <- "output_ftir/plots"
report_dir <- "output_ftir/reports"
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

dir.create(report_dir, recursive = TRUE, showWarnings = FALSE)

# --- Load and clean the data ---
df <- read_csv(data_path)

# --- Filter samples: remove F. Glom and weak carrageenan samples ---
df_filtered <- df  %>%
  filter(!(Iota_Index < 0.3 & Kappa_Index < 0.3 & Lambda_Index < 0.3))

# --- Variables for analysis ---
num_vars <- c("D2S", "DA1070", "DA2S", "DA930", "G2S", "G4S", "S",
              "Kappa_Index", "Iota_Index", "Lambda_Index")

# --- Helper function to run and save PLS-DA analysis ---
run_plsda_analysis <- function(df_input, group_column, ncomp = 3, plot_prefix = "plsda") {
  df_clean <- df_input %>%
    drop_na(all_of(num_vars), !!sym(group_column))

  # Use dplyr::select to avoid masking issues
  X <- df_clean %>%
    dplyr::select(all_of(num_vars)) %>%
    scale() %>%
    as.matrix()

  Y <- as.factor(df_clean[[group_column]])

  # Run PLS-DA
  plsda_res <- plsda(X, Y, ncomp = ncomp)

  # Save PLS-DA individual plot
png(filename = file.path(plot_dir, paste0(plot_prefix, "_", group_column, "_indiv.png")),
    width = 800, height = 600)
plotIndiv(plsda_res, group = Y, legend = TRUE,
          title = paste("PLS-DA:", group_column), ellipse = TRUE)
dev.off()

  # Run cross-validation
  set.seed(123)
  perf_res <- perf(plsda_res, validation = "Mfold", folds = 5, nrepeat = 10, progressBar = TRUE)

  # Save performance plot
  png(filename = file.path(plot_dir, paste0(plot_prefix, "_", group_column, "_perf.png")), width = 800, height = 600)
 plot(perf_res, col = color.mixo(1:3), sd = TRUE)
  dev.off()

  # Save VIP scores
  vip_scores <- map_dfr(1:ncomp, function(i) {
    v <- selectVar(plsda_res, comp = i)$value
    tibble(Variable = rownames(v), VIP = v[,"value.var"], Component = paste0("Comp", i))
  })

  write_csv(vip_scores, file.path(report_dir, paste0(plot_prefix, "_", group_column, "_vip_scores.csv")))

  # Save explained variance bar plot
  explained_var <- plsda_res$prop_expl_var$X[1:ncomp] * 100
  explained_df <- tibble(
    Component = paste0("Comp", 1:ncomp),
    Variance = round(explained_var, 2)
  )

  var_plot <- ggplot(explained_df, aes(x = Component, y = Variance)) +
    geom_col(fill = "steelblue") +
    geom_text(aes(label = paste0(Variance, "%")), vjust = -0.5, size = 4) +
    labs(title = "PLS-DA: Variance Explained by Component",
         y = "Explained Variance (%)", x = "Component") +
    theme_minimal(base_size = 13)

  ggsave(
    filename = file.path(plot_dir, paste0(plot_prefix, "_", group_column, "_explained_variance.png")),
    plot = var_plot, width = 7, height = 5, bg = "white"
  )

  # Return result list
  list(plsda = plsda_res, perf = perf_res)
}

# --- Run analyses ---
# 1. Region-based grouping
res_region <- run_plsda_analysis(df_filtered, group_column = "Location", ncomp = 4, plot_prefix = "region")


# 2. Life_S subset: Tetra vs Carp
df_var_subset <- df_filtered %>%
  filter(Location %in% c("Mendieta", "CC_7H"))

res_life_S <- run_plsda_analysis(df_var_subset, group_column = "variety", ncomp = 4, plot_prefix = "variety")



```
```{r}
```{r non-parametric-carrageenan-inclusive, message=FALSE, warning=FALSE}

# --- Load and Prepare Data ---
# data <- read_csv("your_data.csv")  # Replace with actual data path

# Filter to include all samples and non-NA carrageenan values
df_filtered <- df_filtered %>%
  drop_na(car_per)

# ------------------------
# 1. Kruskal-Wallis Tests
# ------------------------

kruskal_variety <- df_filtered %>%
  kruskal_test(car_per ~ variety)

kruskal_life <- df_filtered %>%
  kruskal_test(car_per ~ Life_S)

# ------------------------
# 2. Pairwise Mann-Whitney U Tests (with Bonferroni Correction)
# ------------------------

pairwise_life <- df_filtered %>%
  pairwise_wilcox_test(car_per ~ Life_S, p.adjust.method = "bonferroni") %>%
  arrange(p.adj)

# ------------------------
# 3. Save Summary Table
# ------------------------

# Save results as CSV (create folder first if needed)
output_dir <- "output/carrageenan_stats"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

write_csv(pairwise_life, file.path(output_dir, "life_stage_pairwise_nonparametric.csv"))

# Display summary in markdown
pairwise_life
```

Kruskal-Wallis Results (all samples):
By Variety: significant — variety continues to influence carrageenan concentration.

By Life Stage: Still shows a significant difference across groups.

Mann-Whitney U Posthoc Results:
Carp vs Unknown again stands out with a low raw p-value (≈ 0.036), though not significant after Bonferroni correction (adjusted p ≈ 0.36).

Other pairwise differences (Carp vs Gam, Gam vs Gam/Tetra, etc.) remain non-significant, but the Carp vs Gam comparison's p-value dropped slightly (now ≈ 0.19), suggesting a trend worth further attention with more data.
### 📚 References
- **Kruskal-Wallis Test**: Kruskal, W. H., & Wallis, W. A. (1952). *Use of ranks in one-criterion variance analysis*. JASA, 47(260), 583–621.
- **Mann-Whitney U Test**: Mann, H. B., & Whitney, D. R. (1947). *On a Test of Whether One of Two Random Variables is Stochastically Larger than the Other*. Ann. Math. Stat., 18(1), 50–60.
- **Bonferroni Correction**: Dunn, O. J. (1961). *Multiple Comparisons Among Means*. JASA, 56(293), 52–64.

```


