---
title: "Carrageenan classification"
author: "Trevor Eakes"
date: "`r Sys.Date()`"
output: html_document
---

Explanation of key steps

Max‐peak extraction (df_peaks): for each sample and each band range, we pull the single highest absorbance.

Arithmetic‐means normalization (df_norm): each band’s peak is divided by the mean of that sample’s peaks, yielding relative intensities around 1.

Index calculation (scores):

Kappa = G4S / DA

Iota = DA2S / DA

Lambda = (G2S + DA2S) / DA

Final tables:

Table3_Intensities: one row per band, showing relative intensities for each sample.

Carrageenan_Scores: per‐sample continuous Kappa, Iota, and Lambda indices.
```{r}
# Required packages


# 1) Read your updated data
df_raw <- read_csv("metaboanalyst_main_input_data.csv")

# 2) Convert to long format (only spectral columns)
df_ftir <- df_raw %>%
  pivot_longer(
    cols = matches("^[0-9]{3,4}$"),  # selects columns like "4000", "3999", etc.
    names_to = "Wavenumber",
    values_to = "Absorbance"
  ) %>%
  mutate(Wavenumber = as.numeric(Wavenumber))

# 2) Define your carrageenan bands
bands <- tribble(
  ~code,     ~name,                                   ~lo,   ~hi,
  "S",       "Sulphate ester (O–SO₃⁻)",              1205,  1265,
  "DA1070",  "C–O band of 3,6-anhydrogalactose",     1055,  1075,
  "DA930",   "3,6-anhydro-D-galactose",              920,   935,
  "G4S",     "D-galactose-4-sulphate",               840,   850,
  "G2S",     "D-galactose-2-sulphate",               825,   830,
  "D2S6S",     "D-galactose-2,6-disulphate",           815,   820,
  "DA2S",    "3,6-anhydro-D-galactose-2-sulphate",   800,   805,
)

# 3) Extract max-peak intensity for each SAMPLE+GROUP & band
df_peaks <- df_ftir %>%
  crossing(bands) %>%
  dplyr::filter(Wavenumber >= lo, Wavenumber <= hi) %>%
  group_by(Sample, code, name, lo, hi) %>%
  slice_max(order_by = Absorbance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(Peak = Absorbance, Peak_Position = Wavenumber)

df_peaks %>%
  group_by(Sample, code) %>%
  summarize(n = n(), .groups = "drop") %>%
  dplyr::filter(n > 1)

#
```

```{r}
# Load required packages


# --- Directories and setup ---
plot_dir <- "output_ftir/plots"
dir_create(plot_dir, recurse = TRUE)

# --- Filter FTIR data to range 1400 to 700 cm⁻¹ ---
df_ftir_subset <- df_ftir %>%
  dplyr::filter(Wavenumber <= 1400, Wavenumber >= 700)

# --- Apply baseline correction to all samples ---
# Step 1: Pivot to wide format (samples × wavenumber matrix)
df_wide <- df_ftir_subset %>%
  mutate(Wavenumber = as.character(Wavenumber)) %>%
  pivot_wider(names_from = Wavenumber, values_from = Absorbance)

sample_ids <- df_wide$Sample
spectra_mat <- df_wide %>% dplyr::select(-Sample)

# Remove columns that could not be converted to numeric (e.g., non-wavenumber names)
spectra_mat <- spectra_mat[, colSums(is.na(sapply(spectra_mat, as.numeric))) == 0]

# Convert to numeric matrix and fix colnames
spectra_mat <- as.matrix(sapply(spectra_mat, as.numeric))
wavenumbers <- as.numeric(colnames(spectra_mat))
colnames(spectra_mat) <- wavenumbers

# Step 2: Apply baseline correction using modpolyfit
baseline_result <- baseline(spectra_mat, method = "modpolyfit", degree = 4)
corrected_mat <- getCorrected(baseline_result)
colnames(corrected_mat) <- wavenumbers

# Step 3: Convert corrected matrix back to long format
corrected_df <- as.data.frame(corrected_mat) %>%
  mutate(Sample = sample_ids) %>%
  pivot_longer(
    cols = -Sample,
    names_to = "Wavenumber",
    values_to = "Absorbance_corrected"
  ) %>%
  mutate(Wavenumber = as.numeric(Wavenumber))

# Step 4: Merge corrected values back into full FTIR data
df_ftir_subset <- df_ftir_subset %>%
  left_join(corrected_df, by = c("Sample", "Wavenumber"))

# --- Filter bands to same range ---
bands_subset <- bands %>%
  dplyr::filter(lo <= 1400, hi >= 700)

# --- Gaussian smoothing function ---
gauss_smooth <- function(x, y, sigma = 3) {
  kernel <- dnorm(-10:10, mean = 0, sd = sigma)
  kernel <- kernel / sum(kernel)
  stats::filter(y, filter = kernel, sides = 2)
}



# --- Apply light smoothing to spectra ---
df_smoothed <- df_ftir_subset %>%
  group_by(Sample) %>%
  arrange(Wavenumber) %>%
  mutate(
    Abs_smoothed = gauss_smooth(Wavenumber, Absorbance),
    is_outlier = if_else(Sample == "Lab_14", "Outlier", "Normal")
  ) %>%
  ungroup() %>%
  drop_na(Abs_smoothed)
#
#----Outlier detection-----
df_wide <- df_ftir_subset %>%
  dplyr::select(Sample, Wavenumber, Absorbance) %>%
  pivot_wider(names_from = Wavenumber, values_from = Absorbance)
# 2. Extract matrix and scale
mat <- df_wide %>% dplyr::select(-Sample) %>% as.matrix()
mat <- scale(mat)

# 3. Run PCA
pca <- prcomp(mat, center = TRUE, scale. = TRUE)
scores <- as.data.frame(pca$x[, 1:5])  # first 5 PCs

# 4. Mahalanobis distance on PCA scores
center <- colMeans(scores)
cov_matrix <- cov(scores)
mahal_dist <- mahalanobis(scores, center, cov_matrix)

# 5. Identify outliers
threshold <- quantile(mahal_dist, 0.95)
outlier_samples <- df_wide$Sample[mahal_dist > threshold]


# --- Generate combined plot ---
# --- Create mean line for legend ---
summary_line <- df_smoothed %>%
  dplyr::group_by(Wavenumber) %>%
  dplyr::summarize(
    mean = mean(Abs_smoothed, na.rm = TRUE),
    sd   = sd(Abs_smoothed, na.rm = TRUE),
    se   = sd / sqrt(dplyr::n()),
    .groups = "drop"
  )
# --- Add 'standard' sample as its own row with styling info ---
standard_row <- df_smoothed %>%
  dplyr::filter(Sample == "standard") %>%
  dplyr::mutate(is_outlier = "Standard", Abs_smoothed = Absorbance) %>%
  dplyr::select(Wavenumber, Abs_smoothed, Sample, is_outlier)

# --- Classify smoothed data as Outlier or Normal ---
df_smoothed <- df_smoothed %>%
  dplyr::mutate(is_outlier = dplyr::if_else(Sample %in% outlier_samples, "Outlier", "Normal"))

# --- Combine all data ---
plot_data_combined <- dplyr::bind_rows(df_smoothed, mean_line, standard_row)

# --- Create base plot ---
combined_plot <- ggplot() +
  geom_ribbon(
    data = summary_line,
    aes(x = Wavenumber, ymin = mean - sd, ymax = mean + se),
    fill = "blue", alpha = 0.3
  ) +
  geom_line(
    data = summary_line,
    aes(x = Wavenumber, y = mean),
    color = "blue", linetype = "solid", linewidth = 1.4
  ) +
  geom_line(
    data = dplyr::filter(plot_data_combined, is_outlier == "Normal"),
    aes(x = Wavenumber, y = Abs_smoothed, group = Sample),
    color = "gray60", linetype = "solid", alpha = 0.35, linewidth = 0.8
  ) +
  geom_line(
    data = dplyr::filter(plot_data_combined, is_outlier == "Outlier"),
    aes(x = Wavenumber, y = Abs_smoothed, group = Sample),
    color = "orangered", linetype = "dashed", alpha = 1, linewidth = 1.0
  ) +
  geom_line(
    data = dplyr::filter(plot_data_combined, is_outlier == "Standard"),
    aes(x = Wavenumber, y = Abs_smoothed, group = Sample),
    color = "forestgreen", linetype = "dashed", alpha = 1, linewidth = 1.2
  ) +
  scale_x_reverse() +
  theme_minimal(base_size = 13) +
  labs(
    title = "FTIR Spectra (700–1400 cm⁻¹, Light Smoothing, with standard and outlier)",
    x = "Wavenumber (cm⁻¹)",
    y = "Absorbance"
  ) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none"  # We'll handle a manual legend if needed
  )

# --- Add shaded carrageenan bands ---
for (i in seq_len(nrow(bands_subset))) {
  band <- bands_subset[i, ]
  center <- (band$lo + band$hi) / 2
  max_height <- max(plot_data_combined$Abs_smoothed, na.rm = TRUE)

  combined_plot <- combined_plot +
    annotate("rect", xmin = band$lo, xmax = band$hi,
             ymin = -Inf, ymax = Inf, alpha = 0.15, fill = "steelblue") +
    annotate("text", x = center, y = max_height * 0.97,
             label = band$code, angle = 90, vjust = -0.5, size = 3.5)
}

# --- Save plot ---
ggsave("output_ftir/plots/all_spectra_light_smoothed2.png",
       plot = combined_plot, width = 9, height = 6, bg = "white")

# --- Apply to all samples ---
unique_samples <- unique(df_ftir_subset$Sample)
sapply(unique_samples, plot_individual_spectrum)

message("\u2705 All-curve and individual FTIR plots saved with light smoothing and shaded bands (700–1400 cm⁻\u00b9 range only).")




```
```{r}
library(ggplot2)
library(cowplot)

# Dummy dataset just for legend creation
legend_df <- data.frame(
  x = 1:4,
  y = 1,
  group = factor(c("Fitted Line", "Normal", "Outlier", "Standard"),
                 levels = c("Fitted Line", "Normal", "Outlier", "Standard"))
)

# Dummy plot to generate the correct legend
legend_plot <- ggplot(legend_df, aes(x = x, y = y, color = group, linetype = group, alpha = group)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c(
    "Fitted Line" = "black",
    "Normal" = "gray30",
    "Outlier" = "orangered",
    "Standard" = "turquoise4"
  )) +
  scale_linetype_manual(values = c(
    "Fitted Line" = "solid",
    "Normal" = "solid",
    "Outlier" = "dashed",
    "Standard" = "dashed"
  )) +
  scale_alpha_manual(values = c(
    "Fitted Line" = 1,
    "Normal" = 0.35,
    "Outlier" = 1,
    "Standard" = 1
  )) +
  guides(
    color = guide_legend(title = "Line Type"),
    linetype = "none",
    alpha = "none"
  ) +
  theme_void() +
  theme(legend.position = "right")

# Extract just the legend grob
legend_grob <- cowplot::get_legend(legend_plot)
#✅ 2. Save the Legend as a Standalone Image

# Save the legend as its own PNG
cowplot::ggsave2(
  filename = "output_ftir/plots/legend_only.png",
  plot = cowplot::plot_grid(legend_grob),
  width = 3.5,
  height = 2.5,
  bg = "white"
)
```


```{r}
# 1. Extract peaks from corrected absorbance
df_peaks <- df_ftir_subset %>%
  crossing(bands) %>%
  dplyr::filter(Wavenumber >= lo, Wavenumber <= hi) %>%
  group_by(Sample, code, name, lo, hi) %>%
  slice_max(order_by = Absorbance_corrected, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(Peak = Absorbance_corrected)

# 2. Pivot to wide format (each peak code = column)
df_peak_wide <- df_peaks %>%
  dplyr::select(Sample, code, Peak) %>%
  pivot_wider(
    names_from = code,
    values_from = Peak
  ) %>%
  mutate(across(-Sample, ~ replace_na(.x, 0)))  # fill missing peaks with 0

# 3. Normalize by arithmetic mean of diagnostic peaks
val_cols <- df_peak_wide %>%
  dplyr::select(where(is.numeric)) %>%
  names()

df_norm <- df_peak_wide %>%
  rowwise() %>%
  mutate(mean_peak = mean(c_across(all_of(val_cols)), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(across(
    all_of(val_cols),
    ~ ifelse(mean_peak > 0, .x / mean_peak, 0)
  )) %>%
  dplyr::select(-mean_peak)

# 4. Load metadata and merge
meta_cols <- c("Sample", "Sample Code", "Location", "region", "variety", "Life_S", "car_per")

metadata_only <- df_ftir_subset %>%
  dplyr::select(all_of(meta_cols)) %>%
  distinct()

final_table <- df_norm %>%
  left_join(metadata_only, by = "Sample") %>%
  relocate(all_of(meta_cols)) %>%
  mutate(
    total_diag   = DA930 + DA2S + G2S + G4S + D2S6S,
    Kappa_Index  = (DA930 + G4S) / total_diag,
    Iota_Index   = (DA930 + DA2S) / total_diag,
    Lambda_Index = (G2S + D2S6S) / total_diag
  ) %>%
  dplyr::select(all_of(meta_cols), all_of(val_cols), Kappa_Index, Iota_Index, Lambda_Index)

# 5. Classify dominant carrageenan type
assign_dominant_type <- function(kappa, iota, lambda, min_score = 0.48, min_gap = 0.07) {
  scores <- c(Kappa = kappa, Iota = iota, Lambda = lambda)
  top_type <- names(which.max(scores))
  top_score <- max(scores)
  other_scores <- scores[names(scores) != top_type]
  closest_gap <- top_score - max(other_scores)

  if (top_score >= min_score && closest_gap >= min_gap) {
    return(top_type)
  } else {
    return("Mixed")
  }
}

final_table$Dominant_Type_New <- pmap_chr(
  final_table[, c("Kappa_Index", "Iota_Index", "Lambda_Index")],
  ~ assign_dominant_type(..1, ..2, ..3)
)

# 6. Save and preview results
write_csv(final_table, "ftir_combined_results.csv")

cat("\n✅ Combined Table Preview:\n")
print(final_table)
```


##Non parametric Carrageenan tests
```{r non-parametric-carrageenan-inclusive, message=FALSE, warning=FALSE}
# --- PCA & Multivariate Analyses ---
# Prepare matrix of peak intensities (normalized or corrected)
data_matrix <- final_table %>%
  dplyr::select(Sample, car_per, Kappa_Index, Iota_Index, Lambda_Index, DA930, DA2S, G2S, G4S, D2S6S) %>%
  column_to_rownames("Sample") %>%
  scale()

# Remove outliers (manually or based on list)
data_matrix <- data_matrix[!rownames(data_matrix) %in% outlier_samples, ]

# Filter final_table
final_table_clean <- final_table %>%
  dplyr::filter(!Sample %in% outlier_samples)

final_table_clean <- final_table_clean %>%
  mutate(Life_S = case_when(
    Life_S %in% c("Tetra", "Gam") ~ "Gam/Tetra",
    TRUE ~ Life_S
  ))

# Extract PCA data (only numeric variables used for PCA)
data_matrix <- final_table_clean %>%
  dplyr::select(Sample, car_per, Kappa_Index, Iota_Index, Lambda_Index) %>%
  column_to_rownames("Sample") %>%
  scale()

# Run PCA
res_pca <- PCA(data_matrix, graph = FALSE)

# Re-join metadata for plotting
meta_pca <- final_table_clean %>%
  dplyr::select(Sample, Life_S, variety)

# Extract PCA coordinates
coords <- as.data.frame(res_pca$ind$coord) %>%
  rownames_to_column("Sample") %>%
  left_join(meta_pca, by = "Sample")

# Plot with ggplot
ggplot(coords, aes(x = Dim.1, y = Dim.2, color = Life_S)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal(base_size = 14) +
  labs(title = "PCA by variety", x = "PC1", y = "PC2") +
  theme(legend.position = "bottom")
# Dendrogram by Life Stage
dist_matrix <- dist(data_matrix)
clust <- hclust(dist_matrix)
plot(clust, labels = meta_pca$variety, main = "Dendrogram by Life Stage")
#
#----Kruskal Walace tests-----#
# List of numeric variables to test
vars_to_test <- c("car_per", "DA930", "DA2S", "G2S", "G4S", "D2S6S", "Kappa_Index", "Iota_Index", "Lambda_Index")

# Kruskal-Wallis by Life_S
kruskal_results_life <- lapply(vars_to_test, function(var) {
  formula <- as.formula(paste(var, "~ Life_S"))
  test <- kruskal.test(formula, data = final_table_clean)
  data.frame(
    Variable = var,
    p_value = test$p.value,
    group = "Life_S"
  )
})

# Kruskal-Wallis by variety
kruskal_results_variety <- lapply(vars_to_test, function(var) {
  formula <- as.formula(paste(var, "~ variety"))
  test <- kruskal.test(formula, data = final_table_clean)
  data.frame(
    Variable = var,
    p_value = test$p.value,
    group = "variety"
  )
})

# Combine results
kruskal_summary <- bind_rows(kruskal_results_life, kruskal_results_variety)

# View or save
print(kruskal_summary)
write.csv(kruskal_summary, "kruskal_summary_ftir_car.csv")
```

Kruskal-Wallis Results (all samples):
By Variety: significant — variety continues to influence carrageenan concentration.

By Life Stage: Still shows a significant difference across groups.

Mann-Whitney U Posthoc Results:
Carp vs Unknown again stands out with a low raw p-value (≈ 0.036), though not significant after Bonferroni correction (adjusted p ≈ 0.36).

Other pairwise differences (Carp vs Gam, Gam vs Gam/Tetra, etc.) remain non-significant, but the Carp vs Gam comparison's p-value dropped slightly (now ≈ 0.19), suggesting a trend worth further attention with more data.
### 📚 References
- **Kruskal-Wallis Test**: Kruskal, W. H., & Wallis, W. A. (1952). *Use of ranks in one-criterion variance analysis*. JASA, 47(260), 583–621.
- **Mann-Whitney U Test**: Mann, H. B., & Whitney, D. R. (1947). *On a Test of Whether One of Two Random Variables is Stochastically Larger than the Other*. Ann. Math. Stat., 18(1), 50–60.
- **Bonferroni Correction**: Dunn, O. J. (1961). *Multiple Comparisons Among Means*. JASA, 56(293), 52–64.

```{r}
# --- Optional: Calibrate kappa index relative to standard ---
kappa_standard_value <- final_table %>%
  dplyr::filter(Sample == "standard") %>%
  pull(Kappa_Index)

if (length(kappa_standard_value) == 1) {
  final_table <- final_table %>%
    mutate(kappa_percent = (Kappa_Index / kappa_standard_value) * 100)

  # --- Classify each sample by dominant carrageenan index ---
  final_table <- final_table %>%
    mutate(Dominant_Type = pmap_chr(
      list(Kappa_Index, Iota_Index, Lambda_Index),
      function(k, i, l) {
        if (is.na(k) | is.na(i) | is.na(l)) return(NA_character_)
        max_val <- max(c(k, i, l))
        if (k == max_val) return("Kappa-rich")
        if (i == max_val) return("Iota-rich")
        if (l == max_val) return("Lambda-rich")
        return(NA_character_)
      }
    ))
} else {
  warning("standard not found or not unique in final_table.")
}


```

#Test baseline calibration
```{r}
df_standard <- df_ftir_subset %>% 
  dplyr::filter(Sample == "standard")
#----baseline cor-----#
# Install if needed
# install.packages("baseline")

library(baseline)

# Convert to matrix format expected by baseline()
mat <- matrix(df_standard$Absorbance, nrow = 1)

# Apply baseline correction (polynomial fit of order 4)
bl <- baseline(mat, method = "modpolyfit", degree = 4)

# Subtract baseline from original
corrected <- getCorrected(bl)[1, ]

# Add to dataframe
df_standard <- df_standard %>%
  mutate(Absorbance_corrected = corrected)

# Redo correction
df_corrected_peaks <- df_standard %>%
  crossing(bands) %>%
  dplyr::filter(Wavenumber >= lo, Wavenumber <= hi) %>%
  group_by(code, name, lo, hi) %>%
  slice_max(order_by = Absorbance_corrected, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::select(code, name, Peak = Absorbance_corrected)
#
#-----Redo peak intensity------#
df_norm <- df_corrected_peaks %>%
  dplyr::select(code, Peak) %>%
  pivot_wider(names_from = code, values_from = Peak)

df_corrected_peaks_wide <- df_corrected_peaks %>%
  pivot_wider(names_from = code, values_from = Peak)
#
df_norm <- df_corrected_peaks_wide %>%
  dplyr::select(-name) %>%       # Drop verbose names
  pivot_longer(cols = everything(), names_to = "code", values_to = "peak") %>%
  drop_na() %>%
  pivot_wider(names_from = code, values_from = peak)
#
df_norm[is.na(df_norm)] <- 0
#
df_scores <- df_norm %>%
  mutate(
    total_diag = DA930 + DA2S + G2S + G4S + D2S6S,
    Kappa_Index  = (DA930 + G4S) / total_diag,
    Iota_Index   = (DA930 + DA2S) / total_diag,
    Lambda_Index = (G2S + D2S6S) / total_diag
  )

#
print("Original Lambda Index:")
print(final_table %>% dplyr::filter(Sample == "standard") %>% pull(Lambda_Index))

print("Corrected Lambda Index:")
print(df_corrected_peaks_wide$Lambda_Index)

```


